Network based multifactorial modelling microRNA-target interaction.

### 2019-09-28 NAR_form_article.Rmd
article NAR form

---
title: "Network based multifactorial modelling microRNA-target interaction."
author:
- Selcen Ari^[Department of Bioengineering, Yildiz Technical University, Istanbul,
  Turkey]
- Alper Yilmaz^[Department of Bioengineering, Yildiz Technical University, Istanbul,
  Turkey]
date: "`r format(Sys.time(), '%d %B %Y')`"
output:
  pdf_document:
    keep_tex: yes
  word_document: default
  html_document: default
link-citations: yes
cls: NAR.cls
template: NAR-sample.tex
bibliography: ref.bib
---

## Abstract

Competing endogenous RNA (ceRNA) regulations and crosstalk between various types of non-coding RNA in human is remarkable in means of miRNA regulation. Many studies have pointed out that an alteration in miRNA:target interaction can result in unexpected changes due to indirect and complex interactions. In this paper, we defined a new network-based model that handles miRNA:ceRNA interactions with expression values. Our model is able to handle miRNA interaction factors such as seed type, binding energy, if provided. Our approach is able to reveal that a perturbation in an element of network affects whole competing elements differently and cooperative efficiencies of miRNAs on common targets could be calculated. Our findings emphasized importance of miRNA:target ratios being crucial, as reported by previous studies. We have showed that the competing elements which have the same or close expression values may not be affected equally from the perturbation because of repression functionality depended on interaction factors of miRNA target pairs. We applied the model to real sample consisting of breast cancer gene and miRNA expression dataset and experimental miRNA:target interaction dataset all generated via high throughput sequencing methods. A gene over-expressed in tumor tissue, namely *SERPINE2*, is used as perturbing element. We have observed that change in expression level of single gene in miRNA:target network is sufficient to perturb regulations in whole network, due to unforeseen and unpredicted regulation which are only visible when considered in network context. Therefore, this model helps unveiling the crosstalk between elements in miRNA:target network where abundance of target and sponge effect are taken into account. The model is scalable and can be plugged in with emerging miRNA effectors such as circRNAs. The model is available as R package [ceRNAnetsim](https://github.com/selcenari/ceRNAnetsim).

## Introduction

@@

As it is possible for miRNAs to suppress multiple targets, an individual mRNA molecule can also be targeted by multiple miRNAs. In that case, the targeted mRNAs exhibit competitor behavior, that is hypothesized as competing endogenous RNAs (ceRNAs) [@ala_integrated_2013; @cesana_deciphering_2013], against their miRNAs. Briefly, Ala et al. have explained the ceRNA hypothesis as disturbance of the other target when one of the targets on a steady-state system that included one miRNA and two target was perturbed with expression change [@ala_integrated_2013]. Regarding interaction between miRNAs and their target in a cell, explaining and predicting results of an individual perturbation is difficult due to complexity of interactions. Various computational and experimental studies have tackled the problem of unraveling ceRNA:miRNA interactions. It has been observed that when abundance of one of the targets of miR-122 was increased, the other targets' expression also slightly increased as a result of decreasing repressive activity of miR-122 on them [@denzler_assessing_2014]. Bosson et al. have developed a mathematical model for changes on total target pool concentration after grouping targets according to affinity and demonstrated that miRNA activity correlated with affinity between miRNA and target [@bosson_endogenous_2014]. Cooperative efficiency of miRNAs as well as competitor behaviors of targets were also studied and it has been demonstrated to be crucial for regulating available mRNA levels of targets [@denzler_impact_2016]. MiRNA:target interactions have been modeled as stoichiometric and catalytic mechanisms and Figliuzzi et al. have recommended handling models in network context [@figliuzzi_micrornas_2013]. The model that can explain miRNA target interaction through topological features has been applied at bipartite network by Nitzan et al. [@nitzan_interactions_2014]. Robinson and Henderson applied the model that handles miRNA:target direct and indirect interactions via common targeting miRNA of genes and target of miRNAs, at bipartite network. It has been demonstrated that all miRNAs and targets in the network can interact with each other through common miRNAs and genes, without interaction between the same type of nodes [@robinson_modelling_2018].Associated genes that are targets of the same miRNAs have been found with help of correlation of gene expression changes in recent algorithm [@markus_list_sponge_2017]. List et al. have specified that their approach can be useful for ceRNA studies and published their approach as an R package.

## Methods
### Construction of miRNA:target network
miRNA and target pairs per line should be provided as edge list to construct the network. At each line minimum required information is expression levels of miRNA and the target. If available, additional data about factors effecting binding or efficiency of miRNA can be provided as separate columns. After construction of the network, amount of miRNA per target is calculated and kept as edge data. Simply, a target will sequester miRNA proportional to its ratio amount among other targets. If additional criteria effecting the binding of miRNA to its target is provided, distribution of miRNA will be calculated accordingly. Target can be mRNA or any other ceRNA (circRNA, ncRNA, etc.) thus, throughout the manuscript terms target, gene and ceRNA are used interchangeably.

### Triggering perturbation and subsequent calculations
Initially, the network is assumed in steady-state (\autoref{fig1}a) condition and needs least one trigger for initiating calculations. The trigger can be a change in expression level of one or more genes (\autoref{fig1}b). After a trigger, the network undergoes iterative cycle of calculations at each of which distribution of miRNA in local neighborhood is recalculated (\autoref{fig1}c). Based on new miRNA distribution, expression level of each node (i.e. ceRNA) is updated. This update results in change of expression value of common target (G4) in system. In this case, the common target acts as trigger for the other group (the targets interacted with M2 miRNA) in the network. Due to common targeted elements, the change in one neighborhood spreads to other neighborhoods (\autoref{fig1}d), consequently have potential to affect whole network due to "ripple effect".

![Schematic presentation of mechanism of network based model. a) In steady state, miRNAs (M: triangles) repress targets (G: circles) according to proportion of target expression level. b) Two fold increase in transcript level of Gene2 (G2) acts as a trigger (shown in red). c) Distribution of miRNA1 (M1) changes. d) The change at expression of common target affects changes of proportional distribution of miRNA2 (M2). Expression values are rounded to integers for simplicity. Gray values on edges indicate initial distribution, red values indicate increase and blue values indicate decrease in distribution. Shades of circles indicate different levels of increase.](Fig1.jpg){#fig1}

@@

The repression efficiency of a miRNA on the individual target ($Eff_{gi}$) is calculated according to equation \eqref{eq:1}; where miRNA expression ($C_m$) in local neighborhood is distributed among targets using individual gene expression levels ($C_{gi}$). For the genes targeted by multiple miRNAs, cooperative activity of miRNAs on a target gene, $R$, is calculated by summing repression activity of each miRNA (Equation \eqref{eq:2}).

\begin{equation} 
    Eff_{gi}= C_m \times C_{gi}/\sum_{1}^{i} C_{gi} \tag{1}\label{eq:1}
\end{equation}

\begin{equation}
   R_{gi}= Eff_{i1} + Eff_{i2} ... \tag{2}\label{eq:2}
\end{equation}

### Multifactorial calculations in miRNA:target network
Interactions between miRNAs and their targets can be affected from various factors. So, our model integrates multiple factors when calculating overall miRNA activity. We classified factors into two categories. Binding factors determine interaction between miRNA and target and they alter amount of miRNA sequestered to target. Efficiency factors dictate degradation efficiency of miRNA on its target and define amount of target to be degraded. In other words, binding factors exert their influence before or during binding, efficiency factors exert their influence after binding. In the literature, binding free energy [@cao_predicting_2012 ; @helwak_mapping_2013] and seed type [@werfel_preferential_2017] in miRNA:target interactions are described as factors effecting binding affinity. Also, binding region (e.g, 5'UTR, 3'UTR, CDS) was shown to drastically affect miRNA degradation efficiency [@hausser_analysis_2013; @helwak_mapping_2013]. Since binding or degradation efficiency from literature have various and broad ranges, both binding and efficiency factors are normalized to their maximum values and scaled to [0,1] interval. The normalized values of factors are used to determine binding activity and miRNA efficiency on targets (\autoref{fig2}). Binding affinities (activity, $Eff$) of miRNAs on each individual gene are calculated as shown in equation \eqref{eq:3}; $C_m$, miRNA expression in the group; $C_{gi}$, individual gene expression; gi, individual gene (\autoref{fig2}c). 
\begin{equation}
Eff_{gi}= C_m \times E^\prime_{gi} \times STE^\prime_{gi} \times C_{gi}/(\sum_{1}^{i} E^\prime_{gi} \times STE^\prime_{gi} \times C_{gi}) \tag{3}\label{eq:3}
\end{equation}

@@

![Calculations to determine of miRNA binding and repression efficiency. $G$, Gene; $M$, miRNA; $STE$, seed type effect; $RE$, Region Effect; $E$, Energy; $STE^\prime$, normalized values of seed type efficiency coefficient; $RE^\prime$, normalized values of region efficiency coefficient; $E^\prime$, normalized values of energy coefficient. Numbering on edges match the pair order in \autoref{tab:one}](fig2.jpg){#fig2}

Not all miRNA:target binding events result in degradation of target. The degradation of target by bound miRNA depends on efficiency factors such as binding region. Exact repression efficiency of miRNA is calculated according to equation \eqref{eq:4} (\autoref{fig2}d); $RE^\prime_{gi}$, normalized values of region efficiency coefficient between miRNA and gene. The cooperative repression activity of miRNAs to their common targets is figured out as shown in \autoref{fig2}e.

### Breast cancer patient dataset

We have applied our model in a real dataset for which experimental measurements of various factors were available. Expression levels of miRNA and genes in tumor and normal tissue of single patient are retrieved from TCGA Research Network: https://www.cancer.gov/tcga. High-throughput experimental datasets which are provided miRNA:gene target pairs with interaction factors [@helwak_mapping_2013; @moore_mirnatarget_2015]. We have combined miRNA and gene expression datasets via miRNA:target gene dataset retrieved from published datasets [@helwak_mapping_2013; @moore_mirnatarget_2015]. Detailed description of network construction and its code is available in Supplementary Materials and Methods. SERPINE2, one of the most effective nodes, has been used as trigger in ceRNA network simulation in breast cancer patient network. Relevant steps and codes are available at  Supplementary Materials and Methods.

@@

## Results and Conclusions
### Network with miRNA:ceRNA expression level
We have developed a network-based approach to assess effects of expression level changes in competitive ceRNA regulation. The basic mode of miRNA repression activity has been based on miRNA and target abundance in various researches [@arvey_target_2010; @denzler_assessing_2014]. Our approach can effortlessly calculate effects of expression changes when abundance levels of miRNAs and targets is only available factor. In sample network given in Figure (\autoref{fig1}), after an increase in expression level of a gene (G2), expression values of other genes also changed due to redistribution of miRNA among its targets. Previous studies have shown that if a gene abundance increases in ceRNA system, expression levels of genes targeted by shared miRNA are also affected [@lai_understanding_2016; @salmena_cerna_2011; @tay_multilayered_2014]. It was observed that expression levels of primary neighborhoods which do not interact with another miRNA of the trigger gene change in relation to their expression. However, changes in expression levels of G4 and G1 differ although they have same initial conditions. This is due to fact that G4 is targeted by two miRNAs, M1 and M2 (\autoref{fig1}d).

@@

![Target regulations with interaction parameters. a) In the steady-state the repression activity of miRNAs on the targets after binding and repression efficiency. b) The changes the repression activities after increasing of G2 expression. c) Perturbation of primary neighborhoods of M1 miRNA (M1 miRNA group). d) Regulation of gene expression of other gene group via triggering target (common target between M1 and M2). ](Fig3.jpg){#fig3}

### ceRNA:Target networks based on interaction factors
Earlier studies reported that miRNA regulatory interactions are affected by different parameters. For example, Xu et al. have investigated the importance of seed pairing type between miRNAs and their targets and target site location by using proteomics dataset [@xu_characterization_2014]. They have proposed that the features of binding between miRNA and target can be critical for miRNA efficiency. In addition, binding energy between miRNA and targets is a significant determinant for miRNA efficiency and it has been reported that strength of miRNA:target interactions is depended on binding energy of complexes [@breda_quantifying_2015]. Similarly, another study revealed that affinity is correlated with seed pairing of miRNA:target pairs and suggested that affinity is correlated with number of canonical seed base pairing [@bosson_endogenous_2014]. Therefore, we have integrated aforementioned interaction parameters that could be useful for predicting miRNA repression activity more accurately. The sample dataset used in \autoref{fig1} is recalculated with additional factors \autoref{tab:one} in effect with same trigger, two fold increase in Gene2 (\autoref{fig3}). When the factors were taken into account in the system, miRNA efficiencies varied as shown in \autoref{fig3}a. Although the miRNA:target expression ratios in steady-state were same in comparison with the sample dataset without factors, efficiency of binding and repression have changed.

@@

When the model run based on expression values of miRNAs and ceRNAs (genes), proportional distribution of miRNA on targets is determined in accordance with \eqref{eq:1, eq:2}. In that case, the system is observed as shown in \autoref{fig1}a at steady-state. However, if interaction factors are taken into account in the approach, specified interaction factors (i.e, binding factors such as seed type and energy) affect distribution of miRNA on targets as calculated in \eqref{eq:3} (shown in \autoref{fig2}c). After distributed miRNA efficiency was re-calculated with degradation factor(s) (i.e region effect) according to \eqref{eq:4}, steady-state condition of network is obtained (shown in \autoref{fig2}d and \autoref{fig3}a). For instance, proportional distribution of G1:M1 interaction in \autoref{fig1}a is calculated lower than the same pair (G1:M1) in the other approach that includes interaction factors (\autoref{fig2}c). But, a decrease is observed in repression efficiency of M1 miRNA on G1 in interaction factor based network approach (\autoref{fig2}d and \autoref{fig3}a).

It is considered that entire miRNAs in the system affect targets according to target:total target ratio \autoref{fig1}a, when interaction factors did not take into account. However, in presence of interaction factors in miRNA:target interaction network (shown in \autoref{tab:one}), miRNAs are distributed according to target:total target ratio and affinity factors in first like shown in \autoref{fig2}c. After affinity mediated proportional distribution of miRNA expression, degradation factor is considered to specify count of repressive miRNA in pairs because entire bound miRNA target pairs might not be resulted with degradation (It is figured out shown in \autoref{fig2}d). For this reason, miRNA1 (M1) has more weak repressive effect (shown as edge variables) on target Gene2 \autoref{fig3}a in comparison with miRNA1 (M1):Gene2 (G2) interaction in \autoref{fig1}a. On the other hand, it is possible that taking into account of these interaction factors could cause increasing miRNA repression activity like in miRNA2:Gene4 interaction \autoref{fig3}a. When expression of Gene2 (G2) increased, expression values of all genes also changed differentially because of contribution of efficiency factors (\autoref{fig3} b,c,d). We have considered that in our approach energy and seed type of pairs is significant for binding and targeted region is important for repression.

\begin{table}[ht]
\centering
\caption{Expression values of elements and interaction factors of miRNA:target interactions in \autoref{fig3}} 
\begin{tabular}{rllrrrrr}
  \hline
 & competing & miRNA & Competing expression & miRNA expression & seed type & region & energy \\ 
  \hline
   & Gene1 & Mir1 & 10000.00 & 1000.00 & 0.43 & 0.30 & -20.00 \\ 
   & Gene2 & Mir1 & 10000.00 & 1000.00 & 0.43 & 0.01 & -15.00 \\ 
   & Gene3 & Mir1 & 5000.00 & 1000.00 & 0.32 & 0.40 & -14.00 \\ 
   & Gene4 & Mir1 & 10000.00 & 1000.00 & 0.23 & 0.50 & -10.00 \\ 
   & Gene4 & Mir2 & 10000.00 & 2000.00 & 0.35 & 0.90 & -12.00 \\ 
   & Gene5 & Mir2 & 5000.00 & 2000.00 & 0.05 & 0.40 & -11.00 \\ 
   & Gene6 & Mir2 & 10000.00 & 2000.00 & 0.01 & 0.80 & -25.00 \\ 
   \hline
\end{tabular}
\label{tab:one}
\end{table}

In regulation of miRNA target sample system (\autoref{fig3}), miRNA (M1) repressive efficiency on the primary triggering gene (G2) is low in steady-state. So it has been observed that the regulatory activities of miRNA on the targets are weak after the increase of miRNA target G2. When model is triggered with two fold increase in expression level of common target (G4), more prominent changes were observed in gene expression levels in network compared to changes caused by G2 increase. Furthermore, change in expression level of target gene that has strong miRNA repression efficiency resulted in evident perturbation in network. On the other hand, it was observed that Gene2 was weakly affected from change of Gene4 expression because of its weak interaction factors. Since perturbation efficiency of each gene is different, we have developed a function which screens each gene in the network for their perturbation efficiencies. **(see details in Supplementary Materials and Methods**. On the other hand, when we applied the method on the minimal dataset, Gene4 has been found to be the most efficient element in terms of number of perturbed elements and miRNA2 (M2) has been found to be causing the highest mean in expression changes.

### Simulation of large and realistic network

@@

The network was finalized by integrating gene and miRNA expression levels from TCGA determined by RNA-Seq. Our  model can successfully simulate perturbations in such large network despite complex behaviors and struggle to reach steady-state **(see details in Supplementary Materials and Methods)**. Simulations show that change in expression level of single gene has potential to affect whole network, perturbing almost all nodes. These observations are in accordance with competing endogenous RNA hypothesis where genes targeted with many common miRNAs subsequently transmit perturbation to neighboring groups.

@@

## Discussion

@@

## Conclusion

@@

## References

>>>>

---
title: Network based multifactorial modelling microRNA-target interaction.
author:
  - name: Selcen Ari
    affiliation: "1"
  - name: Alper Yilmaz
    affiliation: "1"
    correspond: "To whom correspondence should be addressed. Phone: +90 212 383 4627 Fax: +90 212 383 4625  Email: alyilmaz@yildiz.edu.tr"
    separator: "and"
address:
  - code: 1
    address: Department of Bioengineering, Yildiz Technical University, Istanbul, Turkey

abstract: |
  Competing endogenous RNA (ceRNA) regulations and crosstalk between various types of non-coding RNA in human is remarkable in means of miRNA regulation. Many studies have pointed out that an alteration in miRNA:target interaction can result in unexpected changes due to indirect and complex interactions. In this paper, we defined a new network-based model that handles miRNA:ceRNA interactions with expression values. Our model is able to handle miRNA interaction factors such as seed type, binding energy, if provided. Our approach is able to reveal that a perturbation in an element of network affects whole competing elements differently and cooperative efficiencies of miRNAs on common targets could be calculated. Our findings emphasized importance of miRNA:target ratios being crucial, as reported by previous studies. We have showed that the competing elements which have the same or close expression values may not be affected equally from the perturbation because of repression functionality depended on interaction factors of miRNA target pairs. We applied the model to real sample consisting of breast cancer gene and miRNA expression dataset and experimental miRNA:target interaction dataset all generated via high throughput sequencing methods. A gene over-expressed in tumor tissue, namely *SERPINE2*, is used as perturbing element. We have observed that change in expression level of single gene in miRNA:target network is sufficient to perturb regulations in whole network, due to unforeseen and unpredicted regulation which are only visible when considered in network context. Therefore, this model helps unveiling the crosstalk between elements in miRNA:target network where abundance of target and sponge effect are taken into account. The model is scalable and can be plugged in with emerging miRNA effectors such as circRNAs. The model is available as R package ceRNAnetsim https://github.com/selcenari/ceRNAnetsim.

bibliography: NAR-sample.bib
output: rticles::nar_article
header-includes: 
  \usepackage{float} \floatplacement{figure}{H} 
  \usepackage[colorlinks=true,
            linkcolor=blue,
            urlcolor=blue,
            citecolor=blue]{hyperref}
  \usepackage[T1]{fontenc}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```

# INTRODUCTION

@@

As it is possible for miRNAs to suppress multiple targets, an individual mRNA molecule can also be targeted by multiple miRNAs. In that case, the targeted mRNAs exhibit competitor behavior, that is hypothesized as competing endogenous RNAs (ceRNAs) [@ala_integrated_2013; @cesana_deciphering_2013], against their miRNAs. Briefly, Ala et al. have explained the ceRNA hypothesis as disturbance of the other target when one of the targets on a steady-state system that included one miRNA and two target was perturbed with expression change [@ala_integrated_2013]. Regarding interaction between miRNAs and their target in a cell, explaining and predicting results of an individual perturbation is difficult due to complexity of interactions. Various computational and experimental studies have tackled the problem of unraveling ceRNA:miRNA interactions.

\enlargethispage{-65.1pt}

It has been observed that when abundance of one of the targets of miR-122 was increased, the other targets' expression also slightly increased as a result of decreasing repressive activity of miR-122 on them [@denzler_assessing_2014]. Bosson et al. have developed a mathematical model for changes on total target pool concentration after grouping targets according to affinity and demonstrated that miRNA activity correlated with affinity between miRNA and target [@bosson_endogenous_2014]. Cooperative efficiency of miRNAs as well as competitor behaviors of targets were also studied and it has been demonstrated to be crucial for regulating available mRNA levels of targets [@denzler_impact_2016]. MiRNA:target interactions have been modeled as stoichiometric and catalytic mechanisms and Figliuzzi et al. have recommended handling models in network context [@figliuzzi_micrornas_2013]. The model that can explain miRNA target interaction through topological features has been applied at bipartite network by Nitzan et al. [@nitzan_interactions_2014]. Robinson and Henderson applied the model that handles miRNA:target direct and indirect interactions via common targeting miRNA of genes and target of miRNAs, at bipartite network. It has been demonstrated that all miRNAs and targets in the network can interact with each other through common miRNAs and genes, without interaction between the same type of nodes [@robinson_modelling_2018].Associated genes that are targets of the same miRNAs have been found with help of correlation of gene expression changes in recent algorithm [@markus_list_sponge_2017]. List et al. have specified that their approach can be useful for ceRNA studies and published their approach as an R package.

# MATERIALS AND METHODS

## Construction of miRNA:target network
miRNA and target pairs per line should be provided as edge list to construct the network. At each line minimum required information is expression levels of miRNA and the target. If available, additional data about factors effecting binding or efficiency of miRNA can be provided as separate columns. After construction of the network, amount of miRNA per target is calculated and kept as edge data. Simply, a target will sequester miRNA proportional to its ratio amount among other targets. If additional criteria effecting the binding of miRNA to its target is provided, distribution of miRNA will be calculated accordingly. Target can be mRNA or any other ceRNA (circRNA, ncRNA, etc.) thus, throughout the manuscript terms target, gene and ceRNA are used interchangeably.

## Triggering perturbation and subsequent calculations
Initially, the network is assumed in steady-state (Figure \ref{fig:fig1}A) condition and needs least one trigger for initiating calculations. The trigger can be a change in expression level of one or more genes (Figure \ref{fig:fig1}B). After a trigger, the network undergoes iterative cycle of calculations at each of which distribution of miRNA in local neighborhood is recalculated (Figure \ref{fig:fig1}C). Based on new miRNA distribution, expression level of each node (i.e. ceRNA) is updated. This update results in change of expression value of common target (G4) in system. In this case, the common target acts as trigger for the other group (the targets interacted with M2 miRNA) in the network. Due to common targeted elements, the change in one neighborhood spreads to other neighborhoods (Figure \ref{fig:fig1}D), consequently have potential to affect whole network due to "ripple effect".

\begin{figure*}[ht]
\begin{center}
\includegraphics[width=15cm,height=10cm]{Fig1.jpg}
\end{center}
\caption{Schematic presentation of mechanism of network based model. 
    \textbf{A)} In steady state, miRNAs (M: triangles) repress targets (G: circles) according to proportion of target expression level. 
    \textbf{B)} Two fold increase in transcript level of Gene2 (G2) acts as a trigger (shown in red). 
    \textbf{C)} Distribution of miRNA1 (M1) changes. 
    \textbf{D)} The change at expression of common target affects changes of proportional distribution of miRNA2 (M2). Expression values are rounded to integers for simplicity. Gray values on edges indicate initial distribution, red values indicate increase and blue values indicate decrease in distribution. Shades of circles indicate different levels of increase.}
\label{fig:fig1}
\end{figure*}

@@

The repression efficiency of a miRNA on the individual target ($Eff_{gi}$) is calculated according to equation \eqref{eq:1}; where miRNA expression ($C_m$) in local neighborhood is distributed among targets using individual gene expression levels ($C_{gi}$). For the genes targeted by multiple miRNAs, cooperative activity of miRNAs on a target gene, $R$, is calculated by summing repression activity of each miRNA Equation \eqref{eq:2}.
\begin{equation*} 
    Eff_{gi}= C_m \times C_{gi}/\sum_{1}^{i} C_{gi} \tag{1}\label{eq:1}
\end{equation*}

\begin{equation*}
   R_{gi}= Eff_{i1} + Eff_{i2} ... \tag{2}\label{eq:2}
\end{equation*}

## Multifactorial calculations in miRNA:target network
Interactions between miRNAs and their targets can be affected from various factors. So, our model integrates multiple factors when calculating overall miRNA activity. We classified factors into two categories. Binding factors determine interaction between miRNA and target and they alter amount of miRNA sequestered to target. Efficiency factors dictate degradation efficiency of miRNA on its target and define amount of target to be degraded. In other words, binding factors exert their influence before or during binding, efficiency factors exert their influence after binding. In the literature, binding free energy [@cao_predicting_2012 ; @helwak_mapping_2013] and seed type [@werfel_preferential_2017] in miRNA:target interactions are described as factors effecting binding affinity. Also, binding region (e.g, 5'UTR, 3'UTR, CDS) was shown to drastically affect miRNA degradation efficiency [@hausser_analysis_2013; @helwak_mapping_2013]. Since binding or degradation efficiency from literature have various and broad ranges, both binding and efficiency factors are normalized to their maximum values and scaled to [0,1] interval. The normalized values of factors are used to determine binding activity and miRNA efficiency on targets (Figure \ref{fig:fig2}). Binding affinities (activity, $Eff$) of miRNAs on each individual gene are calculated as shown in equation \eqref{eq:3}; $C_m$, miRNA expression in the group; $C_{gi}$, individual gene expression; gi, individual gene (Figure \ref{fig:fig2}C).

\begin{equation}
Eff_{gi}= C_m \times E^\prime_{gi} \times STE^\prime_{gi} \times C_{gi}/(\sum_{1}^{i} E^\prime_{gi} \times STE^\prime_{gi} \times C_{gi}) \tag{3}\label{eq:3}
\end{equation}

@@

\begin{figure}[ht]
\begin{center}
\includegraphics[width=9cm,height=17cm]{Fig2.jpg}
\end{center}
\caption{Calculations to determine of miRNA binding and repression efficiency. $G$, Gene; $M$, miRNA; $STE$, seed type effect; $RE$, Region Effect; $E$, Energy; $STE^\prime$, normalized values of seed type efficiency coefficient; $RE^\prime$, normalized values of region efficiency coefficient; $E^\prime$, normalized values of energy coefficient. Numbering on edges match the pair order in Table \ref{tab:one} }
\label{fig:fig2}
\end{figure}

Not all miRNA:target binding events result in degradation of target. The degradation of target by bound miRNA depends on efficiency factors such as binding region. Exact repression efficiency of miRNA is calculated according to equation \eqref{eq:4} (Figure \ref{fig:fig2}D); $RE^\prime_{gi}$, normalized values of region efficiency coefficient between miRNA and gene. The cooperative repression activity of miRNAs to their common targets is figured out as shown in Figure \ref{fig:fig2}E.

## Breast cancer patient dataset

We have applied our model in a real dataset for which experimental measurements of various factors were available. Expression levels of miRNA and genes in tumor and normal tissue of single patient are retrieved from TCGA Research Network https://www.cancer.gov/tcga. High-throughput experimental datasets which are provided miRNA:gene target pairs with interaction factors [@helwak_mapping_2013; @moore_mirnatarget_2015]. We have combined miRNA and gene expression datasets via miRNA:target gene dataset retrieved from published datasets [@helwak_mapping_2013; @moore_mirnatarget_2015]. Detailed description of network construction and its code is available in Supplementary Materials and Methods. SERPINE2, one of the most effective nodes, has been used as trigger in ceRNA network simulation in breast cancer patient network. Relevant steps and codes are available at  Supplementary Materials and Methods.

@@

# RESULTS

## Network with miRNA:ceRNA expression level
We have developed a network-based approach to assess effects of expression level changes in competitive ceRNA regulation. The basic mode of miRNA repression activity has been based on miRNA and target abundance in various researches [@arvey_target_2010; @denzler_assessing_2014]. Our approach can effortlessly calculate effects of expression changes when abundance levels of miRNAs and targets is only available factor. In sample network given in Figure (\ref{fig:fig1}), after an increase in expression level of a gene (G2), expression values of other genes also changed due to redistribution of miRNA among its targets. Previous studies have shown that if a gene abundance increases in ceRNA system, expression levels of genes targeted by shared miRNA are also affected [@lai_understanding_2016; @salmena_cerna_2011; @tay_multilayered_2014]. It was observed that expression levels of primary neighborhoods which do not interact with another miRNA of the trigger gene change in relation to their expression. However, changes in expression levels of G4 and G1 differ although they have same initial conditions. This is due to fact that G4 is targeted by two miRNAs, M1 and M2 (Figure \ref{fig:fig1}D).

@@

\begin{figure*}[ht]
\begin{center}
\includegraphics[width=15cm,height=8cm]{Fig3.jpg}
\end{center}
\caption{Target regulations with interaction parameters. 
     \textbf{A)} In the steady-state the repression activity of miRNAs on the targets after binding and repression efficiency. 
     \textbf{B)} The changes the repression activities after increasing of G2 expression. 
     \textbf{C)} Perturbation of primary neighborhoods of M1 miRNA (M1 miRNA group). 
     \textbf{D)} Regulation of gene expression of other gene group via triggering target (common target between M1 and M2)}
\label{fig:fig3}
\end{figure*}

## ceRNA:Target networks based on interaction factors
Earlier studies reported that miRNA regulatory interactions are affected by different parameters. For example, Xu et al. have investigated the importance of seed pairing type between miRNAs and their targets and target site location by using proteomics dataset [@xu_characterization_2014]. They have proposed that the features of binding between miRNA and target can be critical for miRNA efficiency. In addition, binding energy between miRNA and targets is a significant determinant for miRNA efficiency and it has been reported that strength of miRNA:target interactions is depended on binding energy of complexes [@breda_quantifying_2015]. Similarly, another study revealed that affinity is correlated with seed pairing of miRNA:target pairs and suggested that affinity is correlated with number of canonical seed base pairing [@bosson_endogenous_2014]. Therefore, we have integrated aforementioned interaction parameters that could be useful for predicting miRNA repression activity more accurately. The sample dataset used in Figure \ref{fig:fig1} is recalculated with additional factors Table \ref{tab:one} in effect with same trigger, two fold increase in Gene2 (Figure \ref{fig:fig3}). When the factors were taken into account in the system, miRNA efficiencies varied as shown in Figure \ref{fig:fig3}A. Although the miRNA:target expression ratios in steady-state were same in comparison with the sample dataset without factors, efficiency of binding and repression have changed.

@@

When the model run based on expression values of miRNAs and ceRNAs (genes), proportional distribution of miRNA on targets is determined in accordance with equations \eqref{eq:1} \eqref{eq:2}. In that case, the system is observed as shown in Figure \ref{fig:fig1}A at steady-state. However, if interaction factors are taken into account in the approach, specified interaction factors (i.e, binding factors such as seed type and energy) affect distribution of miRNA on targets as calculated in equation \eqref{eq:3} (shown in Figure \ref{fig:fig2}C). After distributed miRNA efficiency was re-calculated with degradation factor(s) (i.e region effect) according to equation \eqref{eq:4}, steady-state condition of network is obtained (shown in Figure \ref{fig:fig2}D and Figure \ref{fig:fig3}A). For instance, proportional distribution of G1:M1 interaction in Figure \ref{fig:fig1}A is calculated lower than the same pair (G1:M1) in the other approach that includes interaction factors (Figure \ref{fig:fig2}C). But, a decrease is observed in repression efficiency of M1 miRNA on G1 in interaction factor based network approach (Figure \ref{fig:fig2}D and \ref{fig:fig3}A).

It is considered that entire miRNAs in the system affect targets according to target:total target ratio Figure \ref{fig:fig1}A, when interaction factors did not take into account. However, in presence of interaction factors in miRNA:target interaction network (shown in Table \ref{tab:one}), miRNAs are distributed according to target:total target ratio and affinity factors in first like shown in Figure \ref{fig:fig2}C. After affinity mediated proportional distribution of miRNA expression, degradation factor is considered to specify count of repressive miRNA in pairs because entire bound miRNA target pairs might not be resulted with degradation (It is figured out shown in Figure \ref{fig:fig2}D). For this reason, miRNA1 (M1) has more weak repressive effect (shown as edge variables) on target Gene2 Figure \ref{fig:fig3}A in comparison with miRNA1 (M1):Gene2 (G2) interaction in Figure \ref{fig:fig1}A. On the other hand, it is possible that taking into account of these interaction factors could cause increasing miRNA repression activity like in miRNA2:Gene4 interaction Figure \ref{fig:fig3}A. When expression of Gene2 (G2) increased, expression values of all genes also changed differentially because of contribution of efficiency factors (Figure \ref{fig:fig3} B,C,D). We have considered that in our approach energy and seed type of pairs is significant for binding and targeted region is important for repression.

\begin{table*}[ht]
\centering
\caption{Expression values of elements and interaction factors of miRNA:target interactions} 
\begin{tabular}{rllrrrrr}
  \hline
 & competing & miRNA & Competing expression & miRNA expression & seed type & region & energy \\ 
  \hline
   & Gene1 & Mir1 & 10000.00 & 1000.00 & 0.43 & 0.30 & -20.00 \\ 
   & Gene2 & Mir1 & 10000.00 & 1000.00 & 0.43 & 0.01 & -15.00 \\ 
   & Gene3 & Mir1 & 5000.00 & 1000.00 & 0.32 & 0.40 & -14.00 \\ 
   & Gene4 & Mir1 & 10000.00 & 1000.00 & 0.23 & 0.50 & -10.00 \\ 
   & Gene4 & Mir2 & 10000.00 & 2000.00 & 0.35 & 0.90 & -12.00 \\ 
   & Gene5 & Mir2 & 5000.00 & 2000.00 & 0.05 & 0.40 & -11.00 \\ 
   & Gene6 & Mir2 & 10000.00 & 2000.00 & 0.01 & 0.80 & -25.00 \\ 
   \hline
\end{tabular}
\label{tab:one}
\end{table*}

In regulation of miRNA target sample system (Figure \ref{fig:fig3}), miRNA (M1) repressive efficiency on the primary triggering gene (G2) is low in steady-state. So it has been observed that the regulatory activities of miRNA on the targets are weak after the increase of miRNA target G2. When model is triggered with two fold increase in expression level of common target (G4), more prominent changes were observed in gene expression levels in network compared to changes caused by G2 increase. Furthermore, change in expression level of target gene that has strong miRNA repression efficiency resulted in evident perturbation in network. On the other hand, it was observed that Gene2 was weakly affected from change of Gene4 expression because of its weak interaction factors. Since perturbation efficiency of each gene is different, we have developed a function which screens each gene in the network for their perturbation efficiencies. *(see details in Supplementary Materials and Methods*. On the other hand, when we applied the method on the minimal dataset, Gene4 has been found to be the most efficient element in terms of number of perturbed elements and miRNA2 (M2) has been found to be causing the highest mean in expression changes.

## Simulation of large and realistic network

@@

The network was finalized by integrating gene and miRNA expression levels from TCGA determined by RNA-Seq. Our  model can successfully simulate perturbations in such large network despite complex behaviors and struggle to reach steady-state *(see details in Supplementary Materials and Methods)*. Simulations show that change in expression level of single gene has potential to affect whole network, perturbing almost all nodes. These observations are in accordance with competing endogenous RNA hypothesis where genes targeted with many common miRNAs subsequently transmit perturbation to neighboring groups.

@@

# DISCUSSION

@@

# CONCLUSION

@@

# ACKNOWLEDGEMENTS

Text. Text. Text. Text. Text. Text. Text. Text. Text. Text. Text.
Text. Text. Text. Text.

### Conflict of interest statement.

None declared.

\newpage
